---
title: "ShipStrikeproject"
author: "Andreanne Paul-Chowdhury"
date: "2026-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data provenance

The dataset `Cetacean_shipstrikes_2010Summary.xlsx` was supplied through a
previous university class. Information identifying the original data provider,
sampling design and database documentation was not provided.

The dataset is therefore treated as a historical record-level dataset, and the
analysis does not assume that it represents all cetacean ship strikes or an
unbiased sample of global incidents.


# Introduction

Ship strikes are recognised as a significant threat to marine mammals worldwide. This project investigates patterns in documented cetacean ship strike incidents to identify species most vulnerable to collisions and examine how vessel type and location influence strike severity.

The workflow follows a reproducible data science pipeline consisting of:

- Data import
- Data inspection
- Data cleaning
- Feature engineering
- Exploratory analysis
- Visualisation
- Interpretation

```{r setup, message = FALSE, warning = FALSE}

# Load required packages
library(tidyverse)
library(readxl)
library(janitor)
library(skimr)
library(lubridate)
library(naniar)
library(here)

```

```{r}
# Import the Excel dataset
ship_strikes <- read_excel(
  here("data", "Cetacean_shipstrikes_2010Summary.xlsx")
)

head(ship_strikes)

```


```{r}
#Cleaning column names
ship_strikes <- ship_strikes %>%
  clean_names()
names(ship_strikes)

```

```{r}
#Inspect structure of the dataset
glimpse(ship_strikes)

nrow(ship_strikes) #538 observations

ncol(ship_strikes) # 12 variables

sapply(ship_strikes, class)

summary(ship_strikes)

skim(ship_strikes)

ship_strikes %>%
    summarise(across(everything(),
                     ~sum(is.na(.))))
vis_miss(ship_strikes)
```

```{r}
#looking for duplicates
sum(duplicated(ship_strikes)) #No duplicates
```

```{r}
#Count unique values
#Species
ship_strikes %>%
    count(scientific_name,
          sort = TRUE)
# Vessel type
ship_strikes %>%
    count(vessel_type,
          sort = TRUE)
#Strike outcome
ship_strikes %>%
    count(strike_result, 
          sort = TRUE)
```

```{r}
#Check the date range
range(ship_strikes$evidence_date,
      na.rm = TRUE)
```

```{r}
#Frequency table
# Species
ship_strikes %>%
  count(scientific_name, sort = TRUE)

# Common names
ship_strikes %>%
  count(common_name, sort = TRUE)

# Vessel types
ship_strikes %>%
  count(vessel_type, sort = TRUE)

# Strike outcomes
ship_strikes %>%
  count(strike_result, sort = TRUE)

# Evidence quality
ship_strikes %>%
  count(id_quality, sort = TRUE)

# Ocean regions
ship_strikes %>%
  count(large_area, sort = TRUE)
```

## Data Quality Assessment

Prior to analysis, the dataset was assessed to understand its structure, identify missing values, verify variable types, and detect potential data quality issues. This step ensured that subsequent analyses were based on reliable and appropriately formatted data.

Initial inspection showed the dataset contained 538 observations across 12 variables. Most variables were categorical, with one variable (`evidence_date`) stored as a date. Missing values were concentrated in descriptive fields such as `evidence_date_descriptive`, `collision_evidence`, and `strike_result`, while spatial variables (`large_area` and `small_area`) and `evidence_type` were largely complete. These observations informed the data-cleaning strategy used in the next phase.


# Phase 2: Data Cleaning and Feature Engineering

The purpose of this phase was to standardise inconsistent categories, retain
missing-data information, create variables directly relevant to the research
question, and establish transparent inclusion rules for later analyses.

The original data were preserved unchanged. A separate cleaned dataset was
created so that every transformation remained traceable and reproducible.

The cleaning workflow addressed:

1. character formatting and inconsistent labels;
2. species-identification status;
3. identification confidence;
4. strike-outcome severity;
5. vessel-type categories;
6. ocean-basin categories;
7. temporal precision;
8. analysis-specific inclusion flags.


```{r}
# Preserve the imported dataset
ship_strikes_raw <- ship_strikes

# Confirm that the original dataset has the expected dimensions
stopifnot(
  nrow(ship_strikes_raw) == 538,
  ncol(ship_strikes_raw) == 12
)

glimpse(ship_strikes_raw)
```

```{r}
# Standardise detailed vessel labels
normalise_vessel_type <- function(x) {

  x_clean <- stringr::str_squish(x)
  x_lower <- stringr::str_to_lower(x_clean)

  dplyr::case_when(

    # Missing or explicitly unknown vessels
    is.na(x_clean) |
      x_lower %in% c("not known", "unknown") ~
      "Unknown",

    # Correct three variants of whale-watching vessel
    stringr::str_detect(
      x_lower,
      "whal.*w[h]?atch"
    ) ~
      "Whale-watching vessel",

    # Correct spelling variants for unknown-sized motor yachts
    stringr::str_detect(
      x_lower,
      "motor yacht.*unk"
    ) ~
      "Motor yacht (unknown size)",

    # Correct inflatable-vessel spelling error
    stringr::str_detect(
      x_lower,
      "^rigid"
    ) ~
      "Rigid inflatable/inflatable",

    # Combine Coast Guard spelling and punctuation variants
    stringr::str_detect(
      x_lower,
      "coast guard"
    ) ~
      "US Coast Guard vessel",

    # Standardise case differences
    x_lower == "general cargo" ~ "General cargo",
    x_lower == "oil tanker" ~ "Oil tanker",
    x_lower == "research vessel" ~ "Research vessel",
    x_lower == "cruise ship" ~ "Cruise ship",
    x_lower == "high-speed ferry" ~ "High-speed ferry",

    # Remove spacing difference in gross-tonnage description
    stringr::str_detect(
      x_lower,
      "fishing vessel.*300grt"
    ) ~
      "Fishing vessel (<300 GRT)",

    # Retain any label not addressed above
    TRUE ~ x_clean
  )
}
```

```{r}
# Group detailed vessel descriptions into analytical categories
classify_vessel_type <- function(x) {

  x_lower <- stringr::str_to_lower(x)

  dplyr::case_when(

    x_lower == "unknown" ~
      "Unknown",

    stringr::str_detect(
      x_lower,
      "naval|submarine|coast guard"
    ) ~
      "Government or naval",

    stringr::str_detect(
      x_lower,
      "whale-watching|research vessel"
    ) ~
      "Research or whale-watching",

    stringr::str_detect(
      x_lower,
      "ferry|passenger|cruise|hydrofoil"
    ) ~
      "Passenger or ferry",

    stringr::str_detect(
      x_lower,
      "container|cargo|freighter|oil tanker"
    ) ~
      "Cargo or tanker",

    stringr::str_detect(
      x_lower,
      "fishing|whale catcher"
    ) ~
      "Fishing or whaling",

    stringr::str_detect(
      x_lower,
      paste(
        "motor yacht",
        "sailing yacht",
        "sailing vessel",
        "dinghy",
        "rigid inflatable",
        sep = "|"
      )
    ) ~
      "Recreational or small craft",

    stringr::str_detect(
      x_lower,
      "hopper dredge|pilot boat|other"
    ) ~
      "Other or industrial",

    TRUE ~
      "Other or industrial"
  )
}
```

```{r}
# Clean the dataset
ship_strikes_clean <- ship_strikes_raw |>
  mutate(
    across(
      where(is.character),
      ~ na_if(str_squish(.x), "")
    ),
    record_id = row_number(),
    .before = 1
  )

glimpse(ship_strikes_clean)
```
```{r}
#Clean species information 
ship_strikes_clean <- ship_strikes_clean |>
  mutate(

    # Correct a spelling error without changing the taxonomic meaning
    scientific_name_clean = case_when(
      scientific_name == "Unidentified baleanopterid whale" ~
        "Unidentified balaenopterid whale",

      TRUE ~ scientific_name
    ),

    # Describe the taxonomic resolution of each record
    species_status = case_when(

      is.na(scientific_name_clean) ~
        "Missing scientific name",

      str_detect(
        str_to_lower(scientific_name_clean),
        "^unidentified"
      ) ~
        "Unidentified cetacean",

      scientific_name_clean == "B.m. brevicauda" ~
        "Unresolved abbreviation",

      TRUE ~
        "Identified to species"
    ),

    # Name used specifically for species-level analysis
    species_analysis_name = case_when(

      # This record identifies the species but not the subspecies
      scientific_name_clean ==
        "Balaenoptera acutorostrata ssp." ~
        "Balaenoptera acutorostrata",

      species_status == "Identified to species" ~
        scientific_name_clean,

      TRUE ~
        NA_character_
    )
  )
```

```{r}
ship_strikes_clean |>
  count(
    species_status,
    sort = TRUE,
    name = "number_of_records"
  ) |>
  mutate(
    percentage = round(
      100 * number_of_records / sum(number_of_records),
      1
    )
  )
```

```{r}
#Clean identification quality 
ship_strikes_clean <- ship_strikes_clean |>
  mutate(

    id_quality_clean = case_when(

      is.na(id_quality) ~
        "Missing",

      str_detect(
        str_to_lower(id_quality),
        "^tentative"
      ) ~
        "Tentative",

      TRUE ~ id_quality
    ),

    analysis_confidence = case_when(

      id_quality_clean == "Confirmed" ~
        "Confirmed identification",

      TRUE ~
        "Not confirmed or missing"
    )
  )
```

```{r}
ship_strikes_clean |>
  count(
    id_quality_clean,
    sort = TRUE,
    name = "number_of_records"
  )
```

```{r}
#Clean evidence type 
ship_strikes_clean <- ship_strikes_clean |>
  mutate(

    evidence_type_clean = case_when(

      str_to_lower(evidence_type) == "collision" ~
        "Collision",

      str_to_lower(evidence_type) == "shore" ~
        "Shore",

      str_to_lower(evidence_type) == "at sea" ~
        "At sea",

      str_to_lower(evidence_type) == "found on shore" ~
        "Found on shore",

      str_to_lower(evidence_type) == "found at sea" ~
        "Found at sea",

      str_to_lower(evidence_type) == "stuck on bow" ~
        "Stuck on bow",

      TRUE ~
        "Other"
    ),

    evidence_context = case_when(

      evidence_type_clean == "Collision" ~
        "Reported collision",

      evidence_type_clean %in%
        c("At sea", "Found at sea", "Stuck on bow") ~
        "At-sea evidence",

      evidence_type_clean %in%
        c("Shore", "Found on shore") ~
        "Shore-based evidence",

      TRUE ~
        "Other evidence"
    )
  )
```

```{r}
ship_strikes_clean |>
  count(
    evidence_type,
    evidence_type_clean,
    evidence_context,
    sort = TRUE
  )
```

```{r}
#Clean and classify strike outcome
ship_strikes_clean <- ship_strikes_clean |>
  mutate(

    strike_result_clean = case_when(

      is.na(strike_result) |
        str_to_lower(strike_result) == "not known" ~
        "Unknown",

      str_to_lower(strike_result) == "death" ~
        "Death",

      str_detect(
        str_to_lower(strike_result),
        "severe external visible injury"
      ) ~
        "Severe injury",

      str_detect(
        str_to_lower(strike_result),
        "indeterminate visible injury"
      ) ~
        "Indeterminate injury",

      str_detect(
        str_to_lower(strike_result),
        "apparently minor external injury"
      ) ~
        "Minor injury",

      TRUE ~
        "Other"
    ),

    # Identifies records suitable for outcome analysis
    outcome_known = strike_result_clean != "Unknown",

    # Binary response for fatality-proportion calculations
    fatal_outcome = case_when(
      strike_result_clean == "Death" ~ 1L,
      outcome_known ~ 0L,
      TRUE ~ NA_integer_
    ),

    fatal_label = case_when(
      fatal_outcome == 1L ~ "Fatal",
      fatal_outcome == 0L ~ "Non-fatal known outcome",
      TRUE ~ "Unknown outcome"
    )
  )
```

```{r}
ship_strikes_clean |>
  count(
    strike_result_clean,
    sort = TRUE,
    name = "number_of_records"
  ) |>
  mutate(
    percentage = round(
      100 * number_of_records / sum(number_of_records),
      1
    )
  )
```

```{r}
# Clean and group vessel type
ship_strikes_clean <- ship_strikes_clean |>
  mutate(

    vessel_type_clean =
      normalise_vessel_type(vessel_type),

    vessel_category =
      classify_vessel_type(vessel_type_clean),

    vessel_category = factor(
      vessel_category,
      levels = c(
        "Cargo or tanker",
        "Passenger or ferry",
        "Fishing or whaling",
        "Recreational or small craft",
        "Government or naval",
        "Research or whale-watching",
        "Other or industrial",
        "Unknown"
      )
    )
  )
```

```{r}
ship_strikes_clean |>
  count(
    vessel_category,
    sort = TRUE,
    name = "number_of_records"
  ) |>
  mutate(
    percentage = round(
      100 * number_of_records / sum(number_of_records),
      1
    )
  )

vessel_crosswalk <- ship_strikes_clean |>
  count(
    vessel_type,
    vessel_type_clean,
    vessel_category,
    sort = TRUE,
    name = "number_of_records"
  )

vessel_crosswalk
```

```{r}
#Create ocean-basin categories

ship_strikes_clean <- ship_strikes_clean |>
  mutate(

    large_area_clean = coalesce(
      large_area,
      "Unknown"
    ),

    ocean_basin = case_when(

      str_starts(
        large_area_clean,
        "Atlantic"
      ) ~
        "Atlantic Ocean",

      str_starts(
        large_area_clean,
        "Pacific"
      ) ~
        "Pacific Ocean",

      str_starts(
        large_area_clean,
        "Indian"
      ) ~
        "Indian Ocean",

      str_starts(
        large_area_clean,
        "Arctic"
      ) ~
        "Arctic Ocean",

      str_starts(
        large_area_clean,
        "Southern"
      ) ~
        "Southern Ocean",

      TRUE ~
        "Unknown"
    ),

    ocean_basin = factor(
      ocean_basin,
      levels = c(
        "Atlantic Ocean",
        "Pacific Ocean",
        "Indian Ocean",
        "Southern Ocean",
        "Arctic Ocean",
        "Unknown"
      )
    )
  )
```

```{r}
ship_strikes_clean |>
  count(
    ocean_basin,
    sort = TRUE,
    name = "number_of_records"
  ) |>
  mutate(
    percentage = round(
      100 * number_of_records / sum(number_of_records),
      1
    )
  )

region_crosswalk <- ship_strikes_clean |>
  count(
    large_area_clean,
    ocean_basin,
    sort = TRUE,
    name = "number_of_records"
  )

region_crosswalk
```

```{r}
# ---------------------------------------------------------
# PHASE 2 CHECKPOINT BEFORE DATE CLEANING
# ---------------------------------------------------------

checkpoint_summary <- tibble(
  check = c(
    "Total cleaned records",
    "Identification-quality records",
    "Evidence-type records",
    "Strike-result records",
    "Vessel-category records",
    "Ocean-basin records"
  ),
  total = c(
    nrow(ship_strikes_clean),
    sum(
      ship_strikes_clean |>
        count(id_quality_clean) |>
        pull(n)
    ),
    sum(
      ship_strikes_clean |>
        count(evidence_type_clean) |>
        pull(n)
    ),
    sum(
      ship_strikes_clean |>
        count(strike_result_clean) |>
        pull(n)
    ),
    sum(
      ship_strikes_clean |>
        count(vessel_category) |>
        pull(n)
    ),
    sum(
      ship_strikes_clean |>
        count(ocean_basin) |>
        pull(n)
    )
  )
)

checkpoint_summary
```
```{r}
stopifnot(
  nrow(ship_strikes_clean) == 538,

  sum(
    ship_strikes_clean |>
      count(id_quality_clean) |>
      pull(n)
  ) == 538,

  sum(
    ship_strikes_clean |>
      count(evidence_type_clean) |>
      pull(n)
  ) == 538,

  sum(
    ship_strikes_clean |>
      count(strike_result_clean) |>
      pull(n)
  ) == 538,

  sum(
    ship_strikes_clean |>
      count(vessel_category) |>
      pull(n)
  ) == 538,

  sum(
    ship_strikes_clean |>
      count(ocean_basin) |>
      pull(n)
  ) == 538,

  !anyNA(ship_strikes_clean$strike_result_clean),
  !anyNA(ship_strikes_clean$vessel_category),
  !anyNA(ship_strikes_clean$ocean_basin)
)

message("Checkpoint passed: all 538 records are accounted for.")
```






```{r}
# Construct transparent temporal variables
# ---------------------------------------------------------
# STEP 2.11A: EXTRACT DESCRIPTIVE DATE INFORMATION
# ---------------------------------------------------------

ship_strikes_clean <- ship_strikes_clean |>
  mutate(

    # Standardise the descriptive text for pattern matching
    date_description_clean = str_to_lower(
      str_squish(evidence_date_descriptive)
    ),

    # Extract every four-digit year contained in the description
    extracted_years = str_extract_all(
      coalesce(date_description_clean, ""),
      "(?:18|19|20)\\d{2}"
    ),

    # Count how many years were found
    n_extracted_years = lengths(extracted_years),

    # Extract the first year, where present
    first_extracted_year = map_int(
      extracted_years,
      ~ if (length(.x) == 0) {
        NA_integer_
      } else {
        as.integer(.x[[1]])
      }
    ),

    # Extract the final year, where present
    last_extracted_year = map_int(
      extracted_years,
      ~ if (length(.x) == 0) {
        NA_integer_
      } else {
        as.integer(.x[[length(.x)]])
      }
    ),

    # Detect descriptions such as "1930s" and "late 1980's"
    is_decade_description = str_detect(
      coalesce(date_description_clean, ""),
      "(?:18|19|20)\\d0'?s"
    )
  )
```

```{r}
# ---------------------------------------------------------
# STEP 2.11B: CLASSIFY DATE PRECISION
# ---------------------------------------------------------

ship_strikes_clean <- ship_strikes_clean |>
  mutate(
    date_precision = case_when(

      # Exact date always has the highest priority
      !is.na(evidence_date) ~
        "Exact date",

      # Neither field contains date information
      is.na(evidence_date_descriptive) ~
        "No date information",

      # A reporting date may not represent the incident date
      str_detect(
        coalesce(date_description_clean, ""),
        "\\breport(?:ed|ing)?\\b"
      ) ~
        "Reporting date only",

      # Incident occurred sometime before the stated date
      str_detect(
        coalesce(date_description_clean, ""),
        "\\bbefore\\b"
      ) ~
        "Before specified date",

      # Two or more years indicate an interval
      n_extracted_years >= 2 ~
        "Date range",

      # Examples: "Mid 1930s" or "late 1980's"
      is_decade_description ~
        "Decade only",

      # One identifiable incident year
      n_extracted_years == 1 ~
        "Descriptive year",

      # Text exists but no usable year can be extracted
      TRUE ~
        "Unresolved descriptive date"
    )
  )
```

```{r}
# ---------------------------------------------------------
# STEP 2.11C: CREATE TEMPORAL BOUNDS
# ---------------------------------------------------------

ship_strikes_clean <- ship_strikes_clean |>
  mutate(
    year_start = case_when(

      date_precision == "Exact date" ~
        lubridate::year(evidence_date),

      date_precision == "Descriptive year" ~
        first_extracted_year,

      date_precision == "Date range" ~
        first_extracted_year,

      date_precision == "Decade only" ~
        first_extracted_year,

      TRUE ~
        NA_integer_
    ),

    year_end = case_when(

      date_precision == "Exact date" ~
        lubridate::year(evidence_date),

      date_precision == "Descriptive year" ~
        first_extracted_year,

      date_precision == "Date range" ~
        last_extracted_year,

      date_precision == "Decade only" ~
        first_extracted_year + 9L,

      # This is an upper boundary, not an observed incident year
      date_precision == "Before specified date" ~
        first_extracted_year,

      TRUE ~
        NA_integer_
    )
  )
```

```{r}
# ---------------------------------------------------------
# STEP 2.11D: CREATE ANALYSIS YEAR
# ---------------------------------------------------------

ship_strikes_clean <- ship_strikes_clean |>
  mutate(
    analysis_year = case_when(

      date_precision == "Exact date" ~
        lubridate::year(evidence_date),

      date_precision == "Descriptive year" ~
        first_extracted_year,

      # Ranges, decades and uncertain dates remain missing
      TRUE ~
        NA_integer_
    ),

    year_source = case_when(

      date_precision == "Exact date" ~
        "Exact date",

      date_precision == "Descriptive year" ~
        "Descriptive date",

      TRUE ~
        "No single usable year"
    ),

    analysis_decade = if_else(
      !is.na(analysis_year),
      paste0(
        floor(analysis_year / 10) * 10,
        "s"
      ),
      NA_character_
    )
  )
```

```{r}
ship_strikes_clean <- ship_strikes_clean |>
  select(
    -extracted_years,
    -n_extracted_years,
    -first_extracted_year,
    -last_extracted_year,
    -is_decade_description
  )
```

```{r}
date_precision_summary <- ship_strikes_clean |>
  count(
    date_precision,
    sort = TRUE,
    name = "number_of_records"
  ) |>
  mutate(
    percentage = round(
      100 * number_of_records / sum(number_of_records),
      1
    )
  )

date_precision_summary
```

```{r}
temporal_summary <- ship_strikes_clean |>
  summarise(
    total_records = n(),

    exact_date_records =
      sum(date_precision == "Exact date"),

    descriptive_year_records =
      sum(date_precision == "Descriptive year"),

    usable_analysis_year_records =
      sum(!is.na(analysis_year)),

    records_without_single_year =
      sum(is.na(analysis_year)),

    earliest_analysis_year =
      min(analysis_year, na.rm = TRUE),

    latest_analysis_year =
      max(analysis_year, na.rm = TRUE)
  )

temporal_summary
```

```{r}
# ---------------------------------------------------------
# STEP 2.12: CREATE ANALYSIS-INCLUSION FLAGS
# ---------------------------------------------------------

ship_strikes_clean <- ship_strikes_clean |>
  mutate(

    # Include records that have a usable species name
    include_species_frequency =
      !is.na(species_analysis_name),

    # Include records with a usable species name and known outcome
    include_species_fatality =
      !is.na(species_analysis_name) &
      outcome_known,

    # Sensitivity analysis restricted to confirmed identifications
    include_confirmed_species =
      !is.na(species_analysis_name) &
      id_quality_clean == "Confirmed",

    # Include records with both known outcome and known vessel category
    include_vessel_outcome =
      outcome_known &
      as.character(vessel_category) != "Unknown",

    # Include records with both known outcome and known ocean basin
    include_region_outcome =
      outcome_known &
      as.character(ocean_basin) != "Unknown"
  )
```

```{r}
flag_summary <- tibble(
  measure = c(
    "Total cleaned records",
    "Species-frequency records",
    "Species-fatality records",
    "Confirmed-species records",
    "Vessel-outcome records",
    "Region-outcome records"
  ),
  value = c(
    nrow(ship_strikes_clean),
    sum(ship_strikes_clean$include_species_frequency),
    sum(ship_strikes_clean$include_species_fatality),
    sum(ship_strikes_clean$include_confirmed_species),
    sum(ship_strikes_clean$include_vessel_outcome),
    sum(ship_strikes_clean$include_region_outcome)
  )
)

flag_summary
```

```{r}
stopifnot(
  nrow(ship_strikes_clean) == 538,
  sum(ship_strikes_clean$include_species_frequency) <= 538,
  sum(ship_strikes_clean$include_species_fatality) <= 538,
  sum(ship_strikes_clean$include_confirmed_species) <= 538,
  sum(ship_strikes_clean$include_vessel_outcome) <= 538,
  sum(ship_strikes_clean$include_region_outcome) <= 538
)

message("Step 2.12 passed: analysis-inclusion flags created successfully.")
```
```{r}
# ---------------------------------------------------------
# STEP 2.13: FINAL CLEANING VALIDATION
# ---------------------------------------------------------

stopifnot(

  # All original records were retained
  nrow(ship_strikes_clean) ==
    nrow(ship_strikes_raw),

  # Every record has a unique identifier
  n_distinct(ship_strikes_clean$record_id) ==
    nrow(ship_strikes_clean),

  # Core cleaned categories contain no missing values
  !anyNA(ship_strikes_clean$strike_result_clean),
  !anyNA(ship_strikes_clean$vessel_category),
  !anyNA(ship_strikes_clean$ocean_basin),
  !anyNA(ship_strikes_clean$date_precision),

  # Inclusion flags contain no missing values
  !anyNA(ship_strikes_clean$include_species_frequency),
  !anyNA(ship_strikes_clean$include_species_fatality),
  !anyNA(ship_strikes_clean$include_confirmed_species),
  !anyNA(ship_strikes_clean$include_vessel_outcome),
  !anyNA(ship_strikes_clean$include_region_outcome),

  # Species-fatality records must also be species-frequency records
  all(
    !ship_strikes_clean$include_species_fatality |
      ship_strikes_clean$include_species_frequency
  ),

  # Vessel-outcome records must have known outcomes and vessels
  all(
    !ship_strikes_clean$include_vessel_outcome |
      (
        ship_strikes_clean$outcome_known &
        as.character(
          ship_strikes_clean$vessel_category
        ) != "Unknown"
      )
  ),

  # Region-outcome records must have known outcomes and regions
  all(
    !ship_strikes_clean$include_region_outcome |
      (
        ship_strikes_clean$outcome_known &
        as.character(
          ship_strikes_clean$ocean_basin
        ) != "Unknown"
      )
  ),

  # Exact dates must match their derived analysis years
  all(
    ship_strikes_clean$analysis_year[
      ship_strikes_clean$date_precision == "Exact date"
    ] ==
      lubridate::year(
        ship_strikes_clean$evidence_date[
          ship_strikes_clean$date_precision == "Exact date"
        ]
      )
  )
)

message(
  "Step 2.13 passed: all final cleaning checks were successful."
)
```

```{r}
phase_2_summary <- tibble(
  measure = c(
    "Original records",
    "Cleaned records",
    "Duplicate original records",
    "Species-frequency records",
    "Species-fatality records",
    "Confirmed-species records",
    "Known-outcome records",
    "Documented deaths",
    "Vessel-outcome records",
    "Region-outcome records",
    "Records with a usable analysis year"
  ),

  value = c(
    nrow(ship_strikes_raw),
    nrow(ship_strikes_clean),
    sum(duplicated(ship_strikes_raw)),
    sum(ship_strikes_clean$include_species_frequency),
    sum(ship_strikes_clean$include_species_fatality),
    sum(ship_strikes_clean$include_confirmed_species),
    sum(ship_strikes_clean$outcome_known),
    sum(
      ship_strikes_clean$fatal_outcome == 1,
      na.rm = TRUE
    ),
    sum(ship_strikes_clean$include_vessel_outcome),
    sum(ship_strikes_clean$include_region_outcome),
    sum(!is.na(ship_strikes_clean$analysis_year))
  )
)

phase_2_summary
```

## Analysis-specific datasets

Separate analysis datasets were created from the cleaned master dataset using
the previously defined inclusion flags. This ensured that each analysis used
only records containing the information required for that specific question,
while the complete cleaned dataset remained unchanged.

Creating separate datasets also made changes in sample size transparent and
reduced the risk of applying inconsistent filters during subsequent analyses.

```{r}
# ---------------------------------------------------------
# STEP 2.14: CREATE ANALYSIS-SPECIFIC DATASETS
# ---------------------------------------------------------

# Records suitable for species strike-frequency analysis
species_frequency_data <- ship_strikes_clean |>
  filter(include_species_frequency)

# Records suitable for species fatality-proportion analysis
species_fatality_data <- ship_strikes_clean |>
  filter(include_species_fatality)

# Records suitable for confirmed-identification sensitivity analysis
confirmed_species_data <- ship_strikes_clean |>
  filter(include_confirmed_species)

# Records suitable for vessel category versus outcome analysis
vessel_outcome_data <- ship_strikes_clean |>
  filter(include_vessel_outcome)

# Records suitable for ocean basin versus outcome analysis
region_outcome_data <- ship_strikes_clean |>
  filter(include_region_outcome)

# Records suitable for temporal analysis
temporal_analysis_data <- ship_strikes_clean |>
  filter(!is.na(analysis_year))
```

```{r}
analysis_dataset_summary <- tibble(
  dataset = c(
    "Complete cleaned dataset",
    "Species frequency",
    "Species fatality",
    "Confirmed species",
    "Vessel outcome",
    "Region outcome",
    "Temporal analysis"
  ),

  records = c(
    nrow(ship_strikes_clean),
    nrow(species_frequency_data),
    nrow(species_fatality_data),
    nrow(confirmed_species_data),
    nrow(vessel_outcome_data),
    nrow(region_outcome_data),
    nrow(temporal_analysis_data)
  ),

  percentage_of_cleaned_data = round(
    100 * records / nrow(ship_strikes_clean),
    1
  )
)

analysis_dataset_summary
```
```{r}
stopifnot(

  nrow(species_frequency_data) == 449,
  nrow(species_fatality_data) == 174,
  nrow(confirmed_species_data) == 386,
  nrow(vessel_outcome_data) == 121,
  nrow(region_outcome_data) == 216,
  nrow(temporal_analysis_data) == 526,

  # Species-frequency dataset should contain no missing analysis names
  !anyNA(
    species_frequency_data$species_analysis_name
  ),

  # Species-fatality dataset should contain no missing outcomes
  !anyNA(
    species_fatality_data$fatal_outcome
  ),

  # Confirmed-species dataset should contain confirmed IDs only
  all(
    confirmed_species_data$id_quality_clean ==
      "Confirmed"
  ),

  # Vessel-outcome dataset should not contain unknown vessels
  all(
    as.character(
      vessel_outcome_data$vessel_category
    ) != "Unknown"
  ),

  # Region-outcome dataset should not contain unknown regions
  all(
    as.character(
      region_outcome_data$ocean_basin
    ) != "Unknown"
  ),

  # Temporal dataset should contain no missing analysis years
  !anyNA(
    temporal_analysis_data$analysis_year
  )
)

message(
  "Step 2.14 passed: all analysis datasets were created successfully."
)
```

```{r}
#Save the cleaned data and audit files

dir.create(
  here("data", "processed"),
  recursive = TRUE,
  showWarnings = FALSE
)

saveRDS(
  ship_strikes_clean,
  here(
    "data",
    "processed",
    "cetacean_ship_strikes_clean.rds"
  )
)

write_csv(
  ship_strikes_clean,
  here(
    "data",
    "processed",
    "cetacean_ship_strikes_clean.csv"
  )
)
```

```{r}
# Recreate the date-audit table
date_audit <- ship_strikes_clean |>
  filter(is.na(evidence_date)) |>
  select(
    record_id,
    evidence_date,
    evidence_date_descriptive,
    date_precision,
    year_start,
    year_end,
    analysis_year,
    year_source
  ) |>
  arrange(
    date_precision,
    year_start,
    evidence_date_descriptive
  )

date_audit
```


```{r}
saveRDS(
  species_frequency_data,
  here(
    "data",
    "processed",
    "species_frequency_data.rds"
  )
)

saveRDS(
  species_fatality_data,
  here(
    "data",
    "processed",
    "species_fatality_data.rds"
  )
)

saveRDS(
  confirmed_species_data,
  here(
    "data",
    "processed",
    "confirmed_species_data.rds"
  )
)

saveRDS(
  vessel_outcome_data,
  here(
    "data",
    "processed",
    "vessel_outcome_data.rds"
  )
)

saveRDS(
  region_outcome_data,
  here(
    "data",
    "processed",
    "region_outcome_data.rds"
  )
)

saveRDS(
  temporal_analysis_data,
  here(
    "data",
    "processed",
    "temporal_analysis_data.rds"
  )
)
```

```{r}
write_csv(
  vessel_crosswalk,
  here(
    "data",
    "processed",
    "vessel_cleaning_crosswalk.csv"
  )
)

write_csv(
  region_crosswalk,
  here(
    "data",
    "processed",
    "region_cleaning_crosswalk.csv"
  )
)

write_csv(
  date_audit,
  here(
    "data",
    "processed",
    "date_cleaning_audit.csv"
  )
)

write_csv(
  phase_2_summary,
  here(
    "data",
    "processed",
    "phase_2_summary.csv"
  )
)

write_csv(
  analysis_dataset_summary,
  here(
    "data",
    "processed",
    "analysis_dataset_summary.csv"
  )
)
```

```{r}
list.files(
  here("data", "processed"),
  full.names = FALSE
)
```

```{r}
# Test that saved files can be read successfully
clean_data_check <- readRDS(
  here(
    "data",
    "processed",
    "cetacean_ship_strikes_clean.rds"
  )
)

summary_check <- read_csv(
  here(
    "data",
    "processed",
    "phase_2_summary.csv"
  ),
  show_col_types = FALSE
)

stopifnot(
  nrow(clean_data_check) == 538,
  n_distinct(clean_data_check$record_id) == 538,
  nrow(summary_check) == 11
)

message("Saved Phase 2 files were reopened successfully.")
```

```{r}
glimpse(clean_data_check)
```

# Phase 3: Exploratory Analysis and Results

Phase 3 examined patterns in reported cetacean ship-strike records. Analyses
were conducted using the analysis-specific datasets created during Phase 2.

Species vulnerability was evaluated using two complementary measures:

1. reported strike frequency, defined as the number of documented records for
   each identified species; and
2. fatality proportion, defined as the number of documented deaths divided by
   the number of records with a known outcome.

Reported strike frequency was interpreted as documented strike burden rather
than population-level collision risk because the dataset does not include
species abundance, vessel-traffic exposure or reporting effort.

```{r}
# ---------------------------------------------------------
# STEP 3.1: CREATE SPECIES-LEVEL SUMMARY
# ---------------------------------------------------------

species_summary <- ship_strikes_clean |>
  filter(include_species_frequency) |>
  group_by(species_analysis_name) |>
  summarise(

    # Total documented records for this species
    reported_strikes = n(),

    # Number marked as confirmed identification
    confirmed_identifications = sum(
      id_quality_clean == "Confirmed"
    ),

    # Records with a documented strike outcome
    known_outcomes = sum(
      outcome_known
    ),

    # Records without a documented outcome
    unknown_outcomes = sum(
      !outcome_known
    ),

    # Fatal known outcomes
    documented_deaths = sum(
      fatal_outcome == 1,
      na.rm = TRUE
    ),

    # Non-fatal known outcomes
    non_fatal_known_outcomes = sum(
      fatal_outcome == 0,
      na.rm = TRUE
    ),

    .groups = "drop"
  ) |>
  mutate(

    # Proportion of species records marked confirmed
    confirmed_identification_proportion =
      confirmed_identifications / reported_strikes,

    # Proportion of species records with a known outcome
    outcome_coverage =
      known_outcomes / reported_strikes,

    # Unknown outcomes are excluded from this denominator
    fatality_proportion = if_else(
      known_outcomes > 0,
      documented_deaths / known_outcomes,
      NA_real_
    )
  ) |>
  arrange(
    desc(reported_strikes),
    species_analysis_name
  ) |>
  mutate(
    frequency_rank = row_number()
  )

species_summary
```

```{r}
stopifnot(

  # All species-frequency records are represented
  sum(species_summary$reported_strikes) ==
    nrow(species_frequency_data),

  # All known species outcomes are represented
  sum(species_summary$known_outcomes) ==
    nrow(species_fatality_data),

  # Known outcomes equal fatal plus non-fatal outcomes
  all(
    species_summary$known_outcomes ==
      species_summary$documented_deaths +
      species_summary$non_fatal_known_outcomes
  ),

  # Proportions remain between zero and one
  all(
    species_summary$fatality_proportion >= 0 &
      species_summary$fatality_proportion <= 1,
    na.rm = TRUE
  )
)

message(
  "Step 3.1 passed: species summary created successfully."
)
```

```{r}
nrow(species_summary)
```

```{r}
#Exaxmine reported strike frequency 
# Ten species with the greatest number of documented strikes
top_species_frequency <- species_summary |>
  slice_max(
    order_by = reported_strikes,
    n = 10,
    with_ties = FALSE
  ) |>
  arrange(
    desc(reported_strikes),
    species_analysis_name
  ) |>
  select(
    frequency_rank,
    species_analysis_name,
    reported_strikes,
    confirmed_identifications,
    known_outcomes,
    documented_deaths,
    outcome_coverage,
    fatality_proportion
  )

top_species_frequency
```

```{r}
#Figure 
# Prepare plotting data
species_frequency_plot_data <- top_species_frequency |>
  mutate(
    species_analysis_name = forcats::fct_reorder(
      species_analysis_name,
      reported_strikes
    )
  )

# Create plot
species_frequency_plot <- ggplot(
  species_frequency_plot_data,
  aes(
    x = reported_strikes,
    y = species_analysis_name
  )
) +
  geom_col(
    width = 0.75
  ) +
  geom_text(
    aes(label = reported_strikes),
    hjust = -0.2,
    size = 3.8
  ) +
  scale_x_continuous(
    expand = expansion(
      mult = c(0, 0.12)
    )
  ) +
  labs(
    title = paste0(
      "Ten most frequently identified species; ",
      "n = ",
      nrow(species_frequency_data),
      " species-identified records"
    ),
    x = "Number of documented ship-strike records",
    y = NULL
    
  ) +
  theme_minimal(
    base_size = 12
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(
      face = "bold"
    ),
    plot.caption = element_text(
      hjust = 0
    )
  )

species_frequency_plot
```

```{r}
dir.create(
  here("figures"),
  recursive = TRUE,
  showWarnings = FALSE
)

ggsave(
  filename = here(
    "figures",
    "species_reported_strike_frequency.png"
  ),
  plot = species_frequency_plot,
  width = 9,
  height = 5.5,
  units = "in",
  dpi = 300
)
```


```{r}
#Calculate uncertainty around fatality proportions
# ---------------------------------------------------------
# CUSTOM FUNCTION: WILSON CONFIDENCE INTERVAL
# ---------------------------------------------------------

wilson_interval <- function(
  successes,
  trials,
  confidence_level = 0.95
) {

  # Return missing values when there are no usable trials
  if (
    is.na(trials) ||
    trials == 0
  ) {
    return(
      c(
        lower = NA_real_,
        upper = NA_real_
      )
    )
  }

  z_value <- qnorm(
    1 - (1 - confidence_level) / 2
  )

  observed_proportion <-
    successes / trials

  denominator <-
    1 + (z_value^2 / trials)

  centre <-
    (
      observed_proportion +
        z_value^2 / (2 * trials)
    ) /
    denominator

  margin <-
    (
      z_value *
        sqrt(
          (
            observed_proportion *
              (1 - observed_proportion) /
              trials
          ) +
            (
              z_value^2 /
                (4 * trials^2)
            )
        )
    ) /
    denominator

  c(
    lower = max(
      0,
      centre - margin
    ),
    upper = min(
      1,
      centre + margin
    )
  )
}
```

```{r}
wilson_interval(
  successes = 55,
  trials = 62
)
```
```{r}
#Applying it to every species 
species_summary <- species_summary |>
  mutate(
    fatality_interval = map2(
      documented_deaths,
      known_outcomes,
      ~ wilson_interval(
        successes = .x,
        trials = .y
      )
    ),

    fatality_ci_lower = map_dbl(
      fatality_interval,
      "lower"
    ),

    fatality_ci_upper = map_dbl(
      fatality_interval,
      "upper"
    )
  ) |>
  select(
    -fatality_interval
  )

species_summary
```

```{r}
#3.5 Create a defensible fatality comparison 
#Only retain species that have at least five known outcomes
minimum_known_outcomes <- 5

species_fatality_display <- species_summary |>
  filter(
    known_outcomes >= minimum_known_outcomes
  ) |>
  arrange(
    desc(fatality_proportion),
    desc(known_outcomes)
  ) |>
  select(
    species_analysis_name,
    reported_strikes,
    known_outcomes,
    documented_deaths,
    non_fatal_known_outcomes,
    fatality_proportion,
    fatality_ci_lower,
    fatality_ci_upper,
    outcome_coverage
  )

species_fatality_display
```

```{r}
# Create a folder for analytical result tables
dir.create(
  here("results", "tables"),
  recursive = TRUE,
  showWarnings = FALSE
)
```

```{r}
write_csv(
  species_summary,
  here(
    "results",
    "tables",
    "species_summary.csv"
  )
)
```

```{r}
write_csv(
  top_species_frequency,
  here(
    "results",
    "tables",
    "top_species_frequency.csv"
  )
)
```

```{r}
write_csv(
  species_fatality_display,
  here(
    "results",
    "tables",
    "species_fatality_display.csv"
  )
)
```

```{r}
ggsave(
  filename = here(
    "figures",
    "species_reported_strike_frequency.png"
  ),
  plot = species_frequency_plot,
  width = 9,
  height = 5.5,
  units = "in",
  dpi = 300
)

ggsave(
  filename = here(
    "figures",
    "species_reported_strike_frequency.pdf"
  ),
  plot = species_frequency_plot,
  width = 9,
  height = 5.5,
  units = "in"
)
```

## Species fatality proportions

Fatality proportions were calculated using only records with a known strike
outcome. Unknown outcomes were excluded from the denominator.

To avoid overemphasising percentages based on extremely small samples, the
main figure was restricted to species with at least five known outcomes.
However, all species remained available in the complete species summary.

Wilson 95% confidence intervals were used to display uncertainty around each
fatality proportion.

```{r}
#Calculate overall identified-species fatality proportion
overall_species_fatality <- species_summary |>
  summarise(
    deaths = sum(documented_deaths),
    known_outcomes = sum(known_outcomes),
    fatality_proportion = deaths / known_outcomes
  )

overall_species_fatality
```
```{r}
#Plotting data
species_fatality_plot_data <- species_fatality_display |>
  mutate(
    species_analysis_name = forcats::fct_reorder(
      species_analysis_name,
      fatality_proportion
    ),

    result_label = paste0(
      documented_deaths,
      "/",
      known_outcomes
    )
  )
```

```{r}
# Create plot
species_fatality_plot <- ggplot(
  species_fatality_plot_data,
  aes(
    x = fatality_proportion,
    y = species_analysis_name
  )
) +
  geom_vline(
    xintercept =
      overall_species_fatality$fatality_proportion,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  geom_errorbarh(
    aes(
      xmin = fatality_ci_lower,
      xmax = fatality_ci_upper
    ),
    height = 0.18,
    linewidth = 0.7
  ) +
  geom_point(
    size = 3
  ) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    labels = scales::label_percent(
      accuracy = 1
    ),
    expand = expansion(
      mult = c(0.02, 0.02)
    )
  ) +
  coord_cartesian(
    xlim = c(0, 1),
    clip = "off"
  ) +
  labs(
    title = "Fatality estimates varied among frequently recorded species",
    x = "Fatality proportion among records with known outcomes",
    y = NULL,
  ) +
  theme_minimal(
    base_size = 12
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(
      face = "bold"
    ),
    plot.margin = margin(10,10,10,10)
  )

species_fatality_plot
```

```{r}
ggsave(
  filename = here(
    "figures",
    "species_fatality_proportions.png"
  ),
  plot = species_fatality_plot,
  width = 9,
  height = 5.8,
  units = "in",
  dpi = 300
)

ggsave(
  filename = here(
    "figures",
    "species_fatality_proportions.pdf"
  ),
  plot = species_fatality_plot,
  width = 9,
  height = 5.8,
  units = "in"
)
```

```{r}
#3.7 Confirmed-identification sensitivity analysis
confirmed_frequency_summary <- ship_strikes_clean |>
  filter(include_confirmed_species) |>
  count(
    species_analysis_name,
    name = "confirmed_reported_strikes",
    sort = TRUE
  ) |>
  mutate(
    confirmed_frequency_rank =
      min_rank(
        desc(confirmed_reported_strikes)
      )
  )

#Join full species summary
species_frequency_sensitivity <- species_summary |>
  select(
    species_analysis_name,
    reported_strikes,
    frequency_rank,
    confirmed_identifications
  ) |>
  left_join(
    confirmed_frequency_summary,
    by = "species_analysis_name"
  ) |>
  mutate(
    confirmed_reported_strikes =
      coalesce(
        confirmed_reported_strikes,
        0L
      ),

    rank_change =
      confirmed_frequency_rank -
      frequency_rank
  ) |>
  arrange(
    frequency_rank
  )

species_frequency_sensitivity

#Calculate rank agreement 
frequency_rank_correlation <- cor(
  species_frequency_sensitivity$reported_strikes,
  species_frequency_sensitivity$confirmed_reported_strikes,
  method = "spearman"
)

frequency_rank_correlation

# Compare the top ten lists
top_ten_all <- species_frequency_sensitivity |>
  slice_min(
    frequency_rank,
    n = 10,
    with_ties = FALSE
  ) |>
  pull(species_analysis_name)

top_ten_confirmed <- species_frequency_sensitivity |>
  slice_min(
    confirmed_frequency_rank,
    n = 10,
    with_ties = FALSE
  ) |>
  pull(species_analysis_name)

top_ten_overlap <- length(
  intersect(
    top_ten_all,
    top_ten_confirmed
  )
)

top_ten_overlap
```

```{r}
# 3.8 Confirmed-only fatality comparison 
confirmed_fatality_summary <- ship_strikes_clean |>
  filter(
    include_confirmed_species,
    outcome_known
  ) |>
  group_by(
    species_analysis_name
  ) |>
  summarise(
    confirmed_known_outcomes = n(),

    confirmed_deaths = sum(
      fatal_outcome == 1
    ),

    confirmed_fatality_proportion =
      confirmed_deaths /
      confirmed_known_outcomes,

    .groups = "drop"
  )

#Join with full fatality results
species_fatality_sensitivity <- species_summary |>
  select(
    species_analysis_name,
    known_outcomes,
    documented_deaths,
    fatality_proportion
  ) |>
  left_join(
    confirmed_fatality_summary,
    by = "species_analysis_name"
  ) |>
  mutate(
    fatality_difference =
      confirmed_fatality_proportion -
      fatality_proportion
  ) |>
  arrange(
    desc(known_outcomes)
  )

species_fatality_sensitivity

#Display species with at least 5 outcomes
species_fatality_sensitivity_display <-
  species_fatality_sensitivity |>
  filter(
    known_outcomes >= 5,
    confirmed_known_outcomes >= 5
  )

species_fatality_sensitivity_display

#Summarise sensitivity results
sensitivity_summary <- tibble(
  measure = c(
    "Spearman frequency correlation",
    "Species shared by both top-ten lists",
    "Maximum absolute fatality difference"
  ),

  value = c(
    frequency_rank_correlation,
    top_ten_overlap,
    max(
      abs(
        species_fatality_sensitivity_display$
          fatality_difference
      ),
      na.rm = TRUE
    )
  )
)

sensitivity_summary
```
```{r}
#save new results table 
write_csv(
  species_frequency_sensitivity,
  here(
    "results",
    "tables",
    "species_frequency_sensitivity.csv"
  )
)

write_csv(
  species_fatality_sensitivity,
  here(
    "results",
    "tables",
    "species_fatality_sensitivity.csv"
  )
)

write_csv(
  sensitivity_summary,
  here(
    "results",
    "tables",
    "species_sensitivity_summary.csv"
  )
)
```

## Vessel category and strike outcome

Vessel-outcome analyses were restricted to records containing both a known
vessel category and a known strike outcome. This produced a subset of 121
records.

For each vessel category, the number of known outcomes, documented deaths and
fatality proportion were calculated. Wilson 95% confidence intervals were used
to communicate uncertainty, particularly for vessel categories represented by
small sample sizes.

These results describe patterns within documented records and should not be
interpreted as vessel-specific collision risk because the dataset does not
contain information on vessel abundance, distance travelled or traffic
exposure.

```{r}
# ---------------------------------------------------------
# STEP 3.9: VESSEL CATEGORY AND KNOWN OUTCOMES
# ---------------------------------------------------------

vessel_outcome_summary <- vessel_outcome_data |>
  group_by(vessel_category) |>
  summarise(
    known_outcomes = n(),

    documented_deaths = sum(
      fatal_outcome == 1,
      na.rm = TRUE
    ),

    non_fatal_outcomes = sum(
      fatal_outcome == 0,
      na.rm = TRUE
    ),

    fatality_proportion =
      documented_deaths / known_outcomes,

    .groups = "drop"
  ) |>
  arrange(
    desc(fatality_proportion),
    desc(known_outcomes)
  )

vessel_outcome_summary

#Validate Total
stopifnot(
  sum(vessel_outcome_summary$known_outcomes) ==
    nrow(vessel_outcome_data),

  all(
    vessel_outcome_summary$known_outcomes ==
      vessel_outcome_summary$documented_deaths +
      vessel_outcome_summary$non_fatal_outcomes
  ),

  all(
    between(
      vessel_outcome_summary$fatality_proportion,
      0,
      1
    )
  )
)

message(
  "Vessel outcome totals validated successfully."
)
```

```{r}
#Add Wilson confience intervals
vessel_outcome_summary <- vessel_outcome_summary |>
  mutate(
    fatality_interval = map2(
      documented_deaths,
      known_outcomes,
      ~ wilson_interval(
        successes = .x,
        trials = .y
      )
    ),

    fatality_ci_lower = map_dbl(
      fatality_interval,
      "lower"
    ),

    fatality_ci_upper = map_dbl(
      fatality_interval,
      "upper"
    )
  ) |>
  select(
    -fatality_interval
  )

vessel_outcome_summary
```
```{r}
#Calculate the overall vessel-subset fatality proportion
overall_vessel_fatality <- vessel_outcome_data |>
  summarise(
    known_outcomes = n(),

    documented_deaths = sum(
      fatal_outcome == 1
    ),

    fatality_proportion =
      documented_deaths / known_outcomes
  )

overall_vessel_fatality
```
```{r}
#3.10 Vessel fatality figure 
vessel_plot_data <- vessel_outcome_summary |>
  mutate(
    vessel_category = forcats::fct_reorder(
      vessel_category,
      fatality_proportion
    )
  )

vessel_fatality_plot <- ggplot(
  vessel_plot_data,
  aes(
    x = fatality_proportion,
    y = vessel_category
  )
) +
  geom_vline(
    xintercept =
      overall_vessel_fatality$fatality_proportion,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  geom_errorbarh(
    aes(
      xmin = fatality_ci_lower,
      xmax = fatality_ci_upper
    ),
    height = 0.18,
    linewidth = 0.7
  ) +
  geom_point(
    size = 3
  ) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    labels = scales::label_percent(
      accuracy = 1
    )
  ) +
  coord_cartesian(
    xlim = c(0, 1)
  ) +
  labs(
    title = "Fatality proportions differed among documented vessel categories",
    x = "Fatality proportion among records with known outcomes",
    y = NULL
  ) +
  theme_minimal(
    base_size = 12
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(
      face = "bold"
    )
  )

vessel_fatality_plot
```

```{r}
#Save
ggsave(
  filename = here(
    "figures",
    "vessel_fatality_proportions.png"
  ),
  plot = vessel_fatality_plot,
  width = 9,
  height = 5.5,
  units = "in",
  dpi = 300
)

ggsave(
  filename = here(
    "figures",
    "vessel_fatality_proportions.pdf"
  ),
  plot = vessel_fatality_plot,
  width = 9,
  height = 5.5,
  units = "in"
)
```

```{r}
#3.11 Inspect sample sizes before interpretation
vessel_presentation_table <- vessel_outcome_summary |>
  mutate(
    fatality_percent = scales::percent(
      fatality_proportion,
      accuracy = 0.1
    ),

    outcome_ratio = paste0(
      documented_deaths,
      "/",
      known_outcomes
    )
  ) |>
  select(
    vessel_category,
    known_outcomes,
    documented_deaths,
    outcome_ratio,
    fatality_percent,
    fatality_ci_lower,
    fatality_ci_upper
  )

vessel_presentation_table
```
```{r}
#Save
write_csv(
  vessel_outcome_summary,
  here(
    "results",
    "tables",
    "vessel_outcome_summary.csv"
  )
)

write_csv(
  vessel_presentation_table,
  here(
    "results",
    "tables",
    "vessel_presentation_table.csv"
  )
)
```

```{r}
#3.12 Test the association 
vessel_outcome_contingency <- vessel_outcome_data |>
  count(
    vessel_category,
    fatal_label
  ) |>
  filter(
    fatal_label != "Unknown outcome"
  ) |>
  pivot_wider(
    names_from = fatal_label,
    values_from = n,
    values_fill = 0
  )

vessel_outcome_contingency

#Create the matrix
vessel_test_matrix <- vessel_outcome_data |>
  mutate(
    outcome_group = if_else(
      fatal_outcome == 1,
      "Fatal",
      "Non-fatal"
    )
  ) |>
  count(
    vessel_category,
    outcome_group
  ) |>
  pivot_wider(
    names_from = outcome_group,
    values_from = n,
    values_fill = 0
  ) |>
  column_to_rownames(
    "vessel_category"
  ) |>
  as.matrix()

vessel_test_matrix

#Run the chi-squared test
vessel_chi_test <- chisq.test(
  vessel_test_matrix
)

vessel_chi_test
vessel_chi_test$expected

#Check the expected frequencies
vessel_expected_check <- tibble(
  minimum_expected_count =
    min(vessel_chi_test$expected),

  cells_below_five =
    sum(vessel_chi_test$expected < 5),

  percentage_below_five =
    mean(vessel_chi_test$expected < 5) * 100
)

vessel_expected_check

#28% of cells have expected counts below five, so used a simulated p-value instead
set.seed(5014)

vessel_association_test <- chisq.test(
  vessel_test_matrix,
  simulate.p.value = TRUE,
  B = 10000
)

vessel_association_test
```

```{r}
#save
vessel_test_summary <- tibble(
  test = "Chi-square test with simulated p-value",
  statistic = unname(
    vessel_association_test$statistic
  ),
  degrees_of_freedom = unname(
    vessel_association_test$parameter
  ),
  p_value = vessel_association_test$p.value,
  simulations = 10000
)

vessel_test_summary

write_csv(
  vessel_test_summary,
  here(
    "results",
    "tables",
    "vessel_association_test.csv"
  )
)
```

```{r}
vessel_expected_check <- tibble(
  minimum_expected_count =
    min(vessel_chi_test$expected),

  cells_below_five =
    sum(vessel_chi_test$expected < 5),

  total_cells =
    length(vessel_chi_test$expected),

  percentage_below_five =
    round(
      mean(vessel_chi_test$expected < 5) * 100,
      1
    )
)

vessel_expected_check
```
## Ocean basin and strike outcome

Regional analyses were restricted to records containing both a known strike
outcome and a known ocean basin. This produced a subset of 216 records.

For each ocean basin, the number of known outcomes, documented deaths and
fatality proportion were calculated. Wilson 95% confidence intervals were used
to communicate uncertainty.

These results represent geographic patterns within the documented records.
They should not be interpreted as estimates of absolute regional collision
risk because reporting effort, vessel traffic and cetacean abundance were not
standardised among ocean basins.

```{r}
# ---------------------------------------------------------
# STEP 3.13: OCEAN BASIN AND KNOWN OUTCOMES
# ---------------------------------------------------------

region_outcome_summary <- region_outcome_data |>
  group_by(ocean_basin) |>
  summarise(
    known_outcomes = n(),

    documented_deaths = sum(
      fatal_outcome == 1,
      na.rm = TRUE
    ),

    non_fatal_outcomes = sum(
      fatal_outcome == 0,
      na.rm = TRUE
    ),

    fatality_proportion =
      documented_deaths / known_outcomes,

    .groups = "drop"
  ) |>
  arrange(
    desc(fatality_proportion),
    desc(known_outcomes)
  )

region_outcome_summary

#validate totals
stopifnot(
  sum(region_outcome_summary$known_outcomes) ==
    nrow(region_outcome_data),

  all(
    region_outcome_summary$known_outcomes ==
      region_outcome_summary$documented_deaths +
      region_outcome_summary$non_fatal_outcomes
  ),

  all(
    between(
      region_outcome_summary$fatality_proportion,
      0,
      1
    )
  )
)

message(
  "Regional outcome totals validated successfully."
)

#Add wilson confidence intervals
region_outcome_summary <- region_outcome_summary |>
  mutate(
    fatality_interval = map2(
      documented_deaths,
      known_outcomes,
      ~ wilson_interval(
        successes = .x,
        trials = .y
      )
    ),

    fatality_ci_lower = map_dbl(
      fatality_interval,
      "lower"
    ),

    fatality_ci_upper = map_dbl(
      fatality_interval,
      "upper"
    )
  ) |>
  select(
    -fatality_interval
  )

region_outcome_summary

#Calculate overall fatality proportion
overall_region_fatality <- region_outcome_data |>
  summarise(
    known_outcomes = n(),

    documented_deaths = sum(
      fatal_outcome == 1,
      na.rm = TRUE
    ),

    fatality_proportion =
      documented_deaths / known_outcomes
  )

overall_region_fatality
```
```{r}
#3.14 Region fatality figure
region_plot_data <- region_outcome_summary |>
  mutate(
    ocean_basin = forcats::fct_reorder(
      ocean_basin,
      fatality_proportion
    )
  )
#Create the plot
region_fatality_plot <- ggplot(
  region_plot_data,
  aes(
    x = fatality_proportion,
    y = ocean_basin
  )
) +
  geom_vline(
    xintercept =
      overall_region_fatality$fatality_proportion,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  geom_errorbarh(
    aes(
      xmin = fatality_ci_lower,
      xmax = fatality_ci_upper
    ),
    height = 0.18,
    linewidth = 0.7
  ) +
  geom_point(
    size = 3
  ) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    labels = scales::label_percent(
      accuracy = 1
    )
  ) +
  coord_cartesian(
    xlim = c(0, 1)
  ) +
  labs(
    title = "Fatality proportions varied among documented ocean basins",
    x = "Fatality proportion among records with known outcomes",
    y = NULL
  ) +
  theme_minimal(
    base_size = 12
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(
      face = "bold"
    )
  )

region_fatality_plot
```
```{r}
#save
ggsave(
  filename = here(
    "figures",
    "region_fatality_proportions.png"
  ),
  plot = region_fatality_plot,
  width = 9,
  height = 5.5,
  units = "in",
  dpi = 300
)

ggsave(
  filename = here(
    "figures",
    "region_fatality_proportions.pdf"
  ),
  plot = region_fatality_plot,
  width = 9,
  height = 5.5,
  units = "in"
)
```

```{r}
#3.15 Create region presentation table
region_presentation_table <- region_outcome_summary |>
  mutate(
    outcome_ratio = paste0(
      documented_deaths,
      "/",
      known_outcomes
    ),

    fatality_percent = scales::percent(
      fatality_proportion,
      accuracy = 0.1
    ),

    confidence_interval = paste0(
      scales::percent(
        fatality_ci_lower,
        accuracy = 0.1
      ),
      "–",
      scales::percent(
        fatality_ci_upper,
        accuracy = 0.1
      )
    )
  ) |>
  select(
    ocean_basin,
    known_outcomes,
    documented_deaths,
    non_fatal_outcomes,
    outcome_ratio,
    fatality_percent,
    confidence_interval
  )

region_presentation_table

#save
write_csv(
  region_outcome_summary,
  here(
    "results",
    "tables",
    "region_outcome_summary.csv"
  )
)

write_csv(
  region_presentation_table,
  here(
    "results",
    "tables",
    "region_presentation_table.csv"
  )
)
```
```{r}
#3.16 Test the association
region_test_matrix <- region_outcome_data |>
  mutate(
    outcome_group = if_else(
      fatal_outcome == 1,
      "Fatal",
      "Non-fatal"
    )
  ) |>
  count(
    ocean_basin,
    outcome_group
  ) |>
  pivot_wider(
    names_from = outcome_group,
    values_from = n,
    values_fill = 0
  ) |>
  column_to_rownames(
    "ocean_basin"
  ) |>
  as.matrix()

region_test_matrix

#run the chi-squared test
region_chi_test <- chisq.test(
  region_test_matrix
)

region_chi_test
region_chi_test$expected

#Check the expected frequencies
region_expected_check <- tibble(
  minimum_expected_count =
    min(region_chi_test$expected),

  cells_below_five =
    sum(region_chi_test$expected < 5),

  total_cells =
    length(region_chi_test$expected),

  percentage_below_five =
    round(
      mean(region_chi_test$expected < 5) * 100,
      1
    )
)

region_expected_check

#Run the simulated test 
set.seed(5014)

region_association_test <- chisq.test(
  region_test_matrix,
  simulate.p.value = TRUE,
  B = 10000
)

region_association_test

#p < 0.05: the distribution of fatal and non-fatal outcomes differs among the documented ocean basins.
#Among records with known outcomes, the relative frequency of fatal and non-fatal outcomes differed significantly among ocean basins.
```
```{r}
#save
region_test_summary <- tibble(
  test = "Chi-square test with simulated p-value",

  statistic = unname(
    region_association_test$statistic
  ),

  p_value =
    region_association_test$p.value,

  simulations = 10000
)

region_test_summary

write_csv(
  region_test_summary,
  here(
    "results",
    "tables",
    "region_association_test.csv"
  )
)
```

## Association strength

Cramér's V was calculated alongside the simulated chi-square tests to quantify
the strength of association between strike outcome and each categorical
predictor. Values closer to zero indicate little association, whereas larger
values indicate stronger separation of fatal and non-fatal outcomes among
categories.

Effect sizes were interpreted cautiously because the data represent documented
records rather than a standardised sample of vessel activity or cetacean
exposure.

```{r}
# ---------------------------------------------------------
# STEP 3.17: CALCULATE CRAMER'S V
# ---------------------------------------------------------

calculate_cramers_v <- function(test_object, contingency_matrix) {

  chi_square <- unname(test_object$statistic)
  total_n <- sum(contingency_matrix)
  matrix_dimensions <- dim(contingency_matrix)

  denominator_dimension <- min(
    matrix_dimensions[1] - 1,
    matrix_dimensions[2] - 1
  )

  sqrt(
    chi_square /
      (
        total_n *
          denominator_dimension
      )
  )
}

#calculate both effects sizes
vessel_cramers_v <- calculate_cramers_v(
  test_object = vessel_association_test,
  contingency_matrix = vessel_test_matrix
)

region_cramers_v <- calculate_cramers_v(
  test_object = region_association_test,
  contingency_matrix = region_test_matrix
)

vessel_cramers_v
region_cramers_v

#One combined resutls table
association_summary <- tibble(
  predictor = c(
    "Vessel category",
    "Ocean basin"
  ),

  usable_records = c(
    sum(vessel_test_matrix),
    sum(region_test_matrix)
  ),

  chi_square = c(
    unname(vessel_association_test$statistic),
    unname(region_association_test$statistic)
  ),

  simulated_p_value = c(
    vessel_association_test$p.value,
    region_association_test$p.value
  ),

  cramers_v = c(
    vessel_cramers_v,
    region_cramers_v
  )
) |>
  mutate(
    across(
      c(
        chi_square,
        simulated_p_value,
        cramers_v
      ),
      ~ round(.x, 4)
    )
  )

association_summary

#save
write_csv(
  association_summary,
  here(
    "results",
    "tables",
    "association_summary.csv"
  )
)
```

```{r}
# ---------------------------------------------------------
# STEP 3.18: MODEL FEASIBILITY CHECK
# ---------------------------------------------------------

model_candidate_data <- ship_strikes_clean |>
  filter(
    outcome_known,
    as.character(vessel_category) != "Unknown",
    as.character(ocean_basin) != "Unknown"
  ) |>
  droplevels()

#Check its size
model_candidate_summary <- model_candidate_data |>
  summarise(
    total_records = n(),

    fatal_records = sum(
      fatal_outcome == 1
    ),

    non_fatal_records = sum(
      fatal_outcome == 0
    ),

    vessel_categories =
      n_distinct(vessel_category),

    ocean_basins =
      n_distinct(ocean_basin)
  )

model_candidate_summary

#inspect category sizes
model_candidate_data |>
  count(
    vessel_category,
    sort = TRUE
  )
model_candidate_data |>
  count(
    ocean_basin,
    sort = TRUE
  )
# inspect vessel-by-outcome counts
model_candidate_data |>
  count(
    vessel_category,
    fatal_label
  ) |>
  pivot_wider(
    names_from = fatal_label,
    values_from = n,
    values_fill = 0
  )
#Inspect region-by-outcome counts
model_candidate_data |>
  count(
    ocean_basin,
    fatal_label
  ) |>
  pivot_wider(
    names_from = fatal_label,
    values_from = n,
    values_fill = 0
  )
#Inspect the combined category structure
model_cell_counts <- model_candidate_data |>
  count(
    vessel_category,
    ocean_basin,
    fatal_label
  ) |>
  arrange(
    vessel_category,
    ocean_basin,
    fatal_label
  )

model_cell_counts
#Calculate events-per-parameter diagnostic
number_of_parameters <-
  1 +
  (
    nlevels(
      model_candidate_data$vessel_category
    ) - 1
  ) +
  (
    nlevels(
      model_candidate_data$ocean_basin
    ) - 1
  )

smaller_outcome_group <- min(
  sum(
    model_candidate_data$fatal_outcome == 1
  ),
  sum(
    model_candidate_data$fatal_outcome == 0
  )
)

events_per_parameter <-
  smaller_outcome_group /
  number_of_parameters

model_feasibility_summary <- tibble(
  total_records = nrow(model_candidate_data),
  model_parameters = number_of_parameters,
  smaller_outcome_group = smaller_outcome_group,
  events_per_parameter = round(
    events_per_parameter,
    2
  ),
  empty_vessel_region_cells = sum(
    table(
      model_candidate_data$vessel_category,
      model_candidate_data$ocean_basin
    ) == 0
  )
)

model_feasibility_summary

#Save 
write_csv(
  model_feasibility_summary,
  here(
    "results",
    "tables",
    "model_feasibility_summary.csv"
  )
)
```
## Combined-model feasibility

A combined binomial logistic-regression model was considered to examine vessel
category and ocean basin simultaneously.

The candidate dataset contained 121 complete records and required 10 model
parameters. The smaller outcome group contained 52 records, producing
approximately 5.2 observations per parameter. There were also 11 empty
vessel-category by ocean-basin combinations.

Because of the limited events-per-parameter ratio and sparse category
structure, a combined model was not fitted. Estimated coefficients would be
at risk of instability and could provide misleading apparent precision.

The separate simulated chi-square tests, confidence intervals and effect-size
estimates were therefore retained as the more defensible analytical approach.

```{r}
#3.19 Model decision
model_decision <- model_feasibility_summary |>
  mutate(
    decision = "Combined logistic regression not fitted",

    rationale = paste(
      "Limited events per parameter and 11 empty vessel-region",
      "cells indicated a risk of unstable model estimates."
    )
  )

model_decision

write_csv(
  model_decision,
  here(
    "results",
    "tables",
    "combined_model_decision.csv"
  )
)
```

```{r}
#Step 3.20 Results summary 
# Highest documented species frequency
top_frequency_result <- species_summary |>
  slice_max(
    order_by = reported_strikes,
    n = 1,
    with_ties = FALSE
  )

# Highest species fatality estimate among species with ≥5 outcomes
top_species_fatality_result <- species_fatality_display |>
  slice_max(
    order_by = fatality_proportion,
    n = 1,
    with_ties = FALSE
  )

# Highest vessel estimate among categories with ≥5 outcomes
top_vessel_result <- vessel_outcome_summary |>
  filter(known_outcomes >= 5) |>
  slice_max(
    order_by = fatality_proportion,
    n = 1,
    with_ties = FALSE
  )

# Highest regional estimate among basins with ≥5 outcomes
top_region_result <- region_outcome_summary |>
  filter(known_outcomes >= 5) |>
  slice_max(
    order_by = fatality_proportion,
    n = 1,
    with_ties = FALSE
  )

#Combine them 
phase_3_key_results <- tibble(
  analysis = c(
    "Reported species frequency",
    "Species fatality proportion",
    "Vessel association",
    "Highest vessel-category estimate",
    "Regional association",
    "Highest ocean-basin estimate",
    "Identification sensitivity",
    "Combined-model decision"
  ),

  key_result = c(
    paste0(
      top_frequency_result$species_analysis_name,
      ": ",
      top_frequency_result$reported_strikes,
      " documented records"
    ),

    paste0(
      top_species_fatality_result$species_analysis_name,
      ": ",
      scales::percent(
        top_species_fatality_result$fatality_proportion,
        accuracy = 0.1
      ),
      " (",
      top_species_fatality_result$documented_deaths,
      "/",
      top_species_fatality_result$known_outcomes,
      ")"
    ),

    paste0(
      "Simulated p = ",
      format.pval(
        vessel_association_test$p.value,
        digits = 3,
        eps = 0.001
      ),
      "; Cramer's V = ",
      round(vessel_cramers_v, 2)
    ),

    paste0(
      top_vessel_result$vessel_category,
      ": ",
      scales::percent(
        top_vessel_result$fatality_proportion,
        accuracy = 0.1
      ),
      " (",
      top_vessel_result$documented_deaths,
      "/",
      top_vessel_result$known_outcomes,
      ")"
    ),

    paste0(
      "Simulated p = ",
      format.pval(
        region_association_test$p.value,
        digits = 3,
        eps = 0.001
      ),
      "; Cramer's V = ",
      round(region_cramers_v, 2)
    ),

    paste0(
      top_region_result$ocean_basin,
      ": ",
      scales::percent(
        top_region_result$fatality_proportion,
        accuracy = 0.1
      ),
      " (",
      top_region_result$documented_deaths,
      "/",
      top_region_result$known_outcomes,
      ")"
    ),

    paste0(
      "Frequency correlation = ",
      round(frequency_rank_correlation, 3),
      "; top-ten overlap = ",
      top_ten_overlap,
      "/10"
    ),

    "Full vessel-plus-region logistic model not fitted because of sparse data"
  )
)

phase_3_key_results

#save 
write_csv(
  phase_3_key_results,
  here(
    "results",
    "tables",
    "phase_3_key_results.csv"
  )
)

```
## Phase 3 conclusion

The species analysis showed that documented strike burden and fatality
proportion represented different aspects of apparent vulnerability. Fin and
humpback whales dominated the reported incident records, although their
fatality patterns differed among records with known outcomes.

Restricting the analysis to confirmed identifications produced almost
identical species-frequency rankings, indicating that the principal species
patterns were not driven by uncertain identifications.

Simulated chi-square tests were used to examine associations between fatality
outcome, vessel category and ocean basin because several expected cell counts
were small. Effect sizes were reported using Cramér's V.

A combined logistic-regression model was considered but rejected because the
complete-case dataset contained a relatively low number of observations per
parameter and numerous empty vessel-region combinations. The separate
association analyses were therefore considered more reliable and
interpretable.
